Beyond clines: lineages and haplotype blocks in hybrid zones

### 2016-03-05 drafts/hybrid_zone.tex
YB edited through haplotype length. major changes in P1 of Lineages at selected Loci

\paragraph{Lineages at the selected locus}
Now suppose that the frequency profile of the selected allele, $p(x,t)$, is known.
We seek to first understand how lineages of sampled individuals behave at the selected locus,
from which lineages at linked loci require only a minor modification.

@@

Making the same assumptions as for equation \eqref{eqn:cline_pde}, the derivation in Appendix \ref{apx:lineage_derivation} shows 
how the lineage of an $A$ allele moves as Brownian motion with speed $\sigma$
with bias $\sigma^2 \grad \log p(x,t)$, where $\grad$ is the gradient
(i.e., it moves at speed $\sigma^2$ in the potential $-\log p$).
% \alisa{note to self to read through appendix when written}
To see why this is true, note that the probability that the parent of an $A$ allele found at $x$
lived at position $x-r$ is proportional to $\P\{R=r\} p(x+r)$, where $R$ is the random dispersal distance;
and so the mean displacement from offspring to parent is $\E[R p(x+R)]/\E[p(x+R)]$.
% $$ = \E[ R ( p(x) + R p'(x) ) ]/ p(x) + O(R^3) $$
Using that $\E[R]=0$ and $\E[R^2]=\sigma^2$
and expanding $p()$ to first order about $x$ shows that this mean displacement is approximately $\sigma^2 p'(x)/p(x)$,
which is $\sigma^2$ multiplied by the gradient of $\log p(x)$.

At a geographic position far from the center, $p=1$ to the left, and is proportional to  $\exp(-x\sqrt{s}/\sigma)$ to the right.
Therefore, roughly, lineages on ``their own'' side wander randomly,
while lineages on ``the wrong'' side are pushed at constant speed $\sigma\sqrt{s}$ 
back towards the side where they are more common. 
Since an $A$ allele must by definition have been inherited from the $A$ side of the barrier 
at the time of secondary contact, this push must get more intense the closer it is to the time of secondary contact.
% \alisa{I want to make a figure from the simulations that show the rapid retreat of wrong ancestries back to where they came from}

@@

\paragraph{Lineages at linked loci}
The behavior of a lineage at a linked locus is similar to a selected locus,
except that when in heterozygotes the lineage may move between backgrounds:
if we follow back through time a lineage at a locus we find linked to an $A$ allele, 
it will first tend to be inherited from ancestors to the left (as $A$ lineages drift to the left),
but if it spends time in the hybrid zone, where recombination between selected backgrounds may happen,
it may be found to have been inherited from a $B$-carrying individual,
whose ancestors will tend to be more from the right.

Suppose we sample an allele today at a locus at recombination distance $r$ from the selected site.
We will follow its lineage back through time, using $\tau$ to denote ``generations ago'' to avoid confusion.
If $X_\tau$ is the geographic location of its ancestor $\tau$ generations ago,
and $Z_\tau$ is the identity of the selected allele that ancestor carried,
then we say that $X$ moves as a diffusion pushed by either $\log(p)$ or $\log(1-p)$ 
(just as previously described for a selected allele),
and that $Z$ jumps between $A$ and $B$ at rate either $r (1-p)$  (from $A$ to $B$) or $r p$ (from $B$ to $A$).

>>>>

\paragraph{Lineages at the selected locus}
Now suppose that the frequency profile of the selected allele, $p(x,t)$, is known.
We seek to first understand how lineages of sampled individuals behave at the selected locus. 
These results allow us to extend our analysis to lineages at loci linked to the selected locus.

@@

Making the same assumptions as for equation \eqref{eqn:cline_pde}, the derivation in Appendix \ref{apx:lineage_derivation} shows 
how the lineage of an $A$ allele moves as Brownian motion with speed $\sigma$
with bias $\sigma^2 \grad \log p(x,t)$, where $\grad$ is the gradient
(i.e., it moves at speed $\sigma^2$ in the potential $-\log p$).
% \alisa{note to self to read through appendix when written}
To see why this is true, note that the probability that the parent of an $A$ allele found at $x$
lived at position $x-r$ is proportional to $\P\{R=r\} p(x+r)$, where $R$ is the random dispersal distance;
and so the mean displacement from offspring to parent is $\E[R p(x+R)]/\E[p(x+R)]$.
% $$ = \E[ R ( p(x) + R p'(x) ) ]/ p(x) + O(R^3) $$
Using the fact that $\E[R]=0$ and $\E[R^2]=\sigma^2$
and expanding $p()$ to first order about $x$ shows that this mean displacement is approximately $\sigma^2 p'(x)/p(x)$,
which is $\sigma^2$ multiplied by the gradient of $\log p(x)$.

At a geographic position far from the center, $p=1$ to the left, and is proportional to  $\exp(-x\sqrt{s}/\sigma)$ to the right.
Therefore, roughly, lineages on ``their own'' side wander randomly,
while lineages on ``the wrong'' side are pushed at constant speed $\sigma\sqrt{s}$ 
back towards the side where they are more common. 
Since an $A$ allele must, by definition, have been inherited from the $A$ side of the barrier 
at the time of secondary contact, this push must get more intense the closer it is to the time of secondary contact.
% \alisa{I want to make a figure from the simulations that show the rapid retreat of wrong ancestries back to where they came from}

@@

\paragraph{Lineages at linked loci}
%YB DID SUBSTANTIAL EDITING OF THIS PARAGRAPH
The behavior of a lineage at a linked locus is similar to a selected locus. 
However, there is one important difference -- in heterozygotes the lineage lineage linked to the selected locus may move between backgrounds. 
Therefore, if we follow back through time a lineage at a locus linked to an $A$ allele, 
it will first tend to be inherited from ancestors to the left (as $A$ lineages drift to the left). 
However, with sufficient time in the hybrid zone, recombination allows this linked locus 
to have been inherited from a $B$-carrying individual,
whose ancestors will tend to be more from the right.

Suppose we sample an allele today $r$ Morgans from the selected site. 
We will follow its lineage back through time, using $\tau$ to denote ``generations ago'' to avoid confusion.
If $X_\tau$ is the geographic location of its ancestor $\tau$ generations ago,
and $Z_\tau$ is the identity of the selected allele that ancestor carried,
then we say that $X$ moves as a diffusion pushed by either $\log(p)$ or $\log(1-p)$ 
(as described for a selected allele),
and that $Z$ jumps between $A$ and $B$ at rate either $r (1-p)$  (from $A$ to $B$) or $r p$ (from $B$ to $A$).
